---
title: "Type I Error Rate of a Re-estimation Design"
author: "Gosuke Homma"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Type I Error Rate of a Re-estimation Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r load}
library(bbssr)
```

## Why the rate has to be computed

The size of a test at a fixed sample size is a property of one rejection region. With blinded sample size re-estimation the final sample size depends on the pooled number of interim responders, whose distribution depends on the common response probability $\theta$, and the rejection probability under the null hypothesis averages rejection regions of different sizes with weights that depend on $\theta$. A test that keeps the nominal level at every fixed sample size need not keep it for this average, so the type I error rate of a re-estimation design is computed for the design itself.

Under the null hypothesis $p_{1} = p_{2} = \theta$. Let $n_{11}$ and $n_{12}$ be the interim sample sizes of the two groups, $s$ the pooled number of interim responders, $n_{21}(s)$ and $n_{22}(s)$ the second-stage sample sizes reached from $s$, and $\mathcal{R}_{s}$ the rejection region at the final sample size reached from $s$. The type I error rate is
$$
\mathrm{TIE}(\theta) = \sum_{s = 0}^{n_{11} + n_{12}} \Pr_{\theta}(S = s) \,
\Pr_{\theta}\bigl((X_{1}, X_{2}) \in \mathcal{R}_{s} \mid S = s\bigr) ,
$$
a polynomial in $\theta$ that the package evaluates exactly.

## The rate over the nuisance parameter

`BinaryTypeIErrorBSSR()` evaluates the type I error rate of a re-estimation design and of the fixed-sample design with the initial sample size over a grid of $\theta$, by default `seq(0.005, 0.995, by = 0.005)`, and refines the three largest local maxima on the grid by a one-dimensional optimization. The design below has an assumed difference of 0.3, an initial sample size of 39 per group, an interim analysis after 20 per group, re-estimation by formula (21.3) of Kieser (2020) and the chi-squared test.

```{r tie}
design <- list(Delta.A = 0.3, N1 = 39, N2 = 39, n.interim = c(20, 20), r = 1,
               alpha = 0.025, tar.power = 0.8, Test = 'Chisq', ss.method = 'standard')
tie <- do.call(BinaryTypeIErrorBSSR, design)
tie
plot(tie)
```

```{r tie-max, include = FALSE}
m <- attr(tie, 'max')
```

The largest type I error rate is `r signif(m$TIE[1], 4)` at $\theta =$ `r signif(m$theta[1], 4)` for the re-estimation design and `r signif(m$TIE[2], 4)` at $\theta =$ `r signif(m$theta[2], 4)` for the fixed-sample design. The chi-squared test is not exact, so the fixed-sample design can already exceed the nominal level, and the re-estimation changes the rate again.

## Exact tests in the final analysis

The same design with an exact test in the final analysis gives the following largest rates.

```{r exact-tests}
exact.max <- do.call(rbind, lapply(c('Fisher', 'Boschloo'), function(tst) {
  m <- attr(do.call(BinaryTypeIErrorBSSR, modifyList(design, list(Test = tst))), 'max')
  data.frame(Test = tst, Design = m$Design, theta = signif(m$theta, 4),
             TIE = signif(m$TIE, 4))
}))
exact.max
```

For this rule the largest rate of the re-estimation design is `r exact.max$TIE[exact.max$Test == 'Fisher' & exact.max$Design == 'BSSR']` with Fisher's exact test and `r exact.max$TIE[exact.max$Test == 'Boschloo' & exact.max$Design == 'BSSR']` with the Boschloo test, against the nominal 0.025. Whether the rate stays at or below the level is a property of the rule and the test together. The guarantee of an exact test holds at each fixed sample size and need not carry over to a sample size chosen from the interim data, so the rate has to be checked for each design.

## The adjusted significance level

When the rate exceeds the nominal level, `BinaryAlphaAdjBSSR()` finds the largest nominal level at which the largest type I error rate does not exceed the target level, following Kieser and Friede (2000) and Friede and Kieser (2004). With `adjust = 'test'` (the default) the adjusted level is applied to the final analysis only. The re-estimated sample sizes then do not depend on the level, the largest rate is a non-decreasing step function of it, and the level is found by bisection.

Both searches below use the grid `theta.grid` of the common response probability, which is coarser than the default to keep them short.

```{r adjust-test}
theta.grid <- seq(0.01, 0.99, by = 0.01)
adj <- do.call(BinaryAlphaAdjBSSR, c(design, list(theta = theta.grid)))
adj
```

With `adjust = 'both'` the adjusted level is used in the re-estimation as well. The re-estimated sample sizes then change with the level, the rate need not be monotone in it, and the level is lowered in steps of `step` until the rate is controlled. The search below uses a step of 0.00005, coarser than the default, to keep it short.

```{r adjust-both}
adj.both <- do.call(BinaryAlphaAdjBSSR,
                    c(design, list(adjust = 'both', step = 5e-5, theta = theta.grid)))
adj.both
```

The adjusted level of the re-estimation design is `r signif(adj$alpha.adj[1], 4)` when it applies to the final analysis only and `r signif(adj.both$alpha.adj[1], 4)` when it also applies to the re-estimation. The power at the adjusted level follows from `BinaryPowerBSSR()` with `alpha` set to the adjusted level. The re-estimation keeps the target level through `ss.alpha` when the level was adjusted with `adjust = 'test'`.

```{r adjusted-power}
adjusted <- modifyList(design, list(alpha = adj$alpha.adj[1], ss.alpha = 0.025))
power.adj <- do.call(BinaryPowerBSSR,
                     c(adjusted, list(p = c(0.3, 0.4, 0.5), Delta.T = 0.3)))
power.nominal <- do.call(BinaryPowerBSSR,
                         c(design, list(p = c(0.3, 0.4, 0.5), Delta.T = 0.3)))
data.frame(p = power.adj$p, power.nominal = round(power.nominal$power.BSSR, 4),
           power.adjusted = round(power.adj$power.BSSR, 4))
```

## Conditional rejection probabilities

`BinaryCondRejectBSSR()` takes the calculation apart. Given the pooled number $s$ of interim responders and the pooled number $s_{2}$ of second-stage responders, the number of responders of group 1 in each stage is hypergeometric under the null hypothesis, and the two stages are independent. The rejection probability given $s$ and $s_{2}$,
$$
\mathrm{CRP}(s, s_{2}) = \sum_{x_{11}} \sum_{x_{21}} h_{1}(x_{11} \mid s) \, h_{2}(x_{21} \mid s_{2}) \,
\mathbf{1}\bigl\{(x_{11} + x_{21}, \, s + s_{2} - x_{11} - x_{21}) \in \mathcal{R}_{s}\bigr\} ,
$$
where $h_{1}$ and $h_{2}$ are the hypergeometric probabilities of the two stages, does not depend on $\theta$. The type I error rate is its average over the binomial distributions of $s$ and $s_{2}$,
$$
\mathrm{TIE}(\theta) = \sum_{s} \sum_{s_{2}} b(s; n_{11} + n_{12}, \theta) \,
b\bigl(s_{2}; n_{21}(s) + n_{22}(s), \theta\bigr) \, \mathrm{CRP}(s, s_{2}) ,
$$
where $b$ is the binomial probability. The rate is therefore at most $\alpha$ for every $\theta$ whenever $\mathrm{CRP}$ is at most $\alpha$ for every outcome. The column `CRP.total` is the rejection probability given only the total $s + s_{2}$, with the responder count of group 1 treated as one hypergeometric count among all final patients. Fisher's exact test keeps it at or below $\alpha$ by construction.

```{r crp}
crp <- BinaryCondRejectBSSR(
  Delta.A = 0.3, N1 = 12, N2 = 12, n.interim = c(6, 6), r = 1, alpha = 0.025,
  tar.power = 0.8, Test = 'Fisher', ss.method = 'standard', theta = c(0.3, 0.5)
)
crp
plot(crp)
```

`CRP` exceeds the nominal level at `r attr(crp, 'exceed')[['CRP']]` outcomes and `CRP.total` at `r attr(crp, 'exceed')[['CRP.total']]`. The allocation is fixed within each stage, so the responder count of group 1 given $s$ and $s_{2}$ is the sum of two hypergeometric counts rather than one, and `CRP` differs from `CRP.total` even when the sample size is not re-estimated. Whether the outcomes above the level raise the type I error rate is seen from the decomposition by $s$, which `theta` requests.

```{r by-s}
by.s <- attr(crp, 'by.s')
head(by.s)
attr(crp, 'TIE')
plot(crp, what = 'by.s')
```

`TIE.s` is the type I error rate given $s$, and `contribution` is its product with the probability `prob.s` of $s$. The contributions add up to the type I error rate at each value of `theta`, which is `r paste(signif(attr(crp, 'TIE')$TIE, 4), collapse = ' and ')` here.

## Non-inferiority

With a non-inferiority margin, `BinaryTypeIErrorBSSR()` and `BinaryAlphaAdjBSSR()` evaluate the rate on the boundary of the null hypothesis, parametrized by the pooled response probability, see the `bbssr-non-inferiority` vignette. `BinaryCondRejectBSSR()` covers tests of superiority only, since on the boundary of a non-inferiority hypothesis the two response probabilities differ and the conditional distributions of the responder counts depend on them.

## References

Friede, T. and Kieser, M. (2004). Sample size recalculation for binary data in internal pilot study designs. *Pharmaceutical Statistics*, 3, 269-279.

Kieser, M. (2020). *Methods and Applications of Sample Size Calculation and Recalculation in Clinical Trials*. Springer.

Kieser, M. and Friede, T. (2000). Re-calculating the sample size in internal pilot study designs with control of the type I error rate. *Statistics in Medicine*, 19, 901-911.
