---
title: "Type I Error Rate of a Re-estimation Design"
author: "Gosuke Homma"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Type I Error Rate of a Re-estimation Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r load}
library(bbssr)
```

## Why the rate has to be computed

The size of a test at a fixed sample size is a property of one rejection region. With blinded sample size re-estimation the final sample size depends on the pooled number of interim responders, whose distribution depends on the common response probability $\theta$, and the rejection probability under the null hypothesis averages rejection regions of different sizes with weights that depend on $\theta$. A test that keeps the nominal level at every fixed sample size need not keep it for this average, so the type I error rate of a re-estimation design is computed for the design itself.

Under the null hypothesis $p_{1} = p_{2} = \theta$. Let $n_{11}$ and $n_{12}$ be the interim sample sizes of the two groups, $s$ the pooled number of interim responders, $n_{21}(s)$ and $n_{22}(s)$ the second-stage sample sizes reached from $s$, and $\mathcal{R}_{s}$ the rejection region at the final sample size reached from $s$. The type I error rate is
$$
\mathrm{TIE}(\theta) = \sum_{s = 0}^{n_{11} + n_{12}} \Pr_{\theta}(S = s) \,
\Pr_{\theta}\bigl((X_{1}, X_{2}) \in \mathcal{R}_{s} \mid S = s\bigr) ,
$$
a polynomial in $\theta$ that the package evaluates exactly.

## The rate over the nuisance parameter

`BinaryTypeIErrorBSSR()` evaluates the type I error rate of a re-estimation design and of the fixed-sample design with the initial sample size over a grid of $\theta$, by default `seq(0, 1, by = 0.005)`. The largest value is not read off the grid. With the default `maximize = 'certified'` the function finds the largest value over the whole interval spanned by the grid, here $[0, 1]$, together with an upper bound that the rate does not exceed anywhere in that interval. The `bbssr-statistical-methods` vignette describes the method. The design below has an assumed difference of 0.3, an initial sample size of 39 per group, an interim analysis after 20 per group, re-estimation by formula (21.3) of Kieser (2020) and the chi-squared test.

```{r tie}
design <- list(Delta.A = 0.3, N1 = 39, N2 = 39, n.interim = c(20, 20), r = 1,
               alpha = 0.025, tar.power = 0.8, Test = 'Chisq', ss.method = 'standard')
tie <- do.call(BinaryTypeIErrorBSSR, design)
tie
attr(tie, 'max')
plot(tie)
```

```{r tie-max, include = FALSE}
m <- attr(tie, 'max')
stopifnot(nrow(m) == 2, all(m$bound >= m$TIE), all(m$bound - m$TIE <= 1e-12 + 1e-15))
```

The largest type I error rate is `r signif(m$TIE[1], 4)` at $\theta =$ `r signif(m$theta[1], 4)` for the re-estimation design and `r signif(m$TIE[2], 4)` at $\theta =$ `r signif(m$theta[2], 4)` for the fixed-sample design. The column `bound` of the attribute `max` exceeds these values by at most $10^{-12}$, so the largest rates are known to that accuracy whatever the spacing of the grid. The chi-squared test is not exact, so the fixed-sample design can already exceed the nominal level, and the re-estimation changes the rate again.

## Exact tests in the final analysis

The same design with an exact test in the final analysis gives the following largest rates.

```{r exact-tests}
exact.max <- do.call(rbind, lapply(c('Fisher', 'Boschloo'), function(tst) {
  m <- attr(do.call(BinaryTypeIErrorBSSR, modifyList(design, list(Test = tst))), 'max')
  data.frame(Test = tst, Design = m$Design, theta = signif(m$theta, 4),
             TIE = signif(m$TIE, 4))
}))
exact.max
```

For this rule the largest rate of the re-estimation design is `r exact.max$TIE[exact.max$Test == 'Fisher' & exact.max$Design == 'BSSR']` with Fisher's exact test and `r exact.max$TIE[exact.max$Test == 'Boschloo' & exact.max$Design == 'BSSR']` with the Boschloo test, against the nominal 0.025. Whether the rate stays at or below the level is a property of the rule and the test together. The guarantee of an exact test holds at each fixed sample size and need not carry over to a sample size chosen from the interim data, so the rate has to be checked for each design.

## The adjusted significance level

When the rate exceeds the nominal level, `BinaryAlphaAdjBSSR()` lowers the nominal level until the largest type I error rate does not exceed the target level, following Kieser and Friede (2000) and Friede and Kieser (2004). With `adjust = 'test'` (the default) the adjusted level is applied to the final analysis only. The re-estimated sample sizes then do not depend on the level, the largest rate is a non-decreasing step function of it, and the level is found by bisection. Every level between the largest p-value that is rejected and the smallest p-value that is not gives the same rejection regions, and the reported level is the largest value with at most six significant digits in that range, so it can be used as printed.

Both searches below use the grid `theta.grid` from 0.01 to 0.99 in steps of 0.01, which is coarser than the default to keep them short. The grid serves the search: a level is assessed on the grid, and with the default `maximize = 'certified'` the level found is accepted only if the upper bound of the type I error rate over the whole interval $[0.01, 0.99]$ does not exceed the target level. The columns `max.TIE.bound` and `max.TIE.adj.bound` of the result hold the bounds at the nominal and at the adjusted level.

```{r adjust-test}
theta.grid <- seq(0.01, 0.99, by = 0.01)
adj <- do.call(BinaryAlphaAdjBSSR, c(design, list(theta = theta.grid)))
adj
```

With `adjust = 'both'` the adjusted level is used in the re-estimation as well. The re-estimated sample sizes then change with the level, the rate need not be monotone in it, and the level is lowered in steps of `step` until the rate is controlled. The result is the first level on this grid of steps at which the rate is controlled. The levels that control the rate need not form an interval, so the result can change with the step. The published values of Friede and Kieser (2004, Section 5) and Kieser (2020, Example 21.1) are reproduced only when the re-estimation also uses the adjusted level, as the script `reproduce-published.R` shows (see the `bbssr-validation` vignette). The search below uses a step of 0.00005, coarser than the default, to keep it short.

```{r adjust-both}
adj.both <- do.call(BinaryAlphaAdjBSSR,
                    c(design, list(adjust = 'both', step = 5e-5, theta = theta.grid)))
adj.both
```

The adjusted level of the re-estimation design is `r adj$alpha.adj[1]` when it applies to the final analysis only and `r adj.both$alpha.adj[1]` when it also applies to the re-estimation. At these levels the upper bounds of the type I error rate over $[0.01, 0.99]$ are `r signif(adj$max.TIE.adj.bound[1], 4)` and `r signif(adj.both$max.TIE.adj.bound[1], 4)`. The power at the adjusted level follows from `BinaryPowerBSSR()` with `alpha` set to the adjusted level. The re-estimation keeps the target level through `ss.alpha` when the level was adjusted with `adjust = 'test'`.

```{r adjusted-power}
adjusted <- modifyList(design, list(alpha = adj$alpha.adj[1], ss.alpha = 0.025))
power.adj <- do.call(BinaryPowerBSSR,
                     c(adjusted, list(p = c(0.3, 0.4, 0.5), Delta.T = 0.3)))
power.nominal <- do.call(BinaryPowerBSSR,
                         c(design, list(p = c(0.3, 0.4, 0.5), Delta.T = 0.3)))
data.frame(p = power.adj$p, power.nominal = round(power.nominal$power.BSSR, 4),
           power.adjusted = round(power.adj$power.BSSR, 4))
```

## Conditional rejection probabilities

`BinaryCondRejectBSSR()` takes the calculation apart. Given the pooled number $s$ of interim responders and the pooled number $s_{2}$ of second-stage responders, the number of responders of group 1 in each stage is hypergeometric under the null hypothesis, and the two stages are independent. The rejection probability given $s$ and $s_{2}$,
$$
\mathrm{CRP}(s, s_{2}) = \sum_{x_{11}} \sum_{x_{21}} h_{1}(x_{11} \mid s) \, h_{2}(x_{21} \mid s_{2}) \,
\mathbf{1}\bigl\{(x_{11} + x_{21}, \, s + s_{2} - x_{11} - x_{21}) \in \mathcal{R}_{s}\bigr\} ,
$$
where $h_{1}$ and $h_{2}$ are the hypergeometric probabilities of the two stages, does not depend on $\theta$. The type I error rate is its average over the binomial distributions of $s$ and $s_{2}$,
$$
\mathrm{TIE}(\theta) = \sum_{s} \sum_{s_{2}} b(s; n_{11} + n_{12}, \theta) \,
b\bigl(s_{2}; n_{21}(s) + n_{22}(s), \theta\bigr) \, \mathrm{CRP}(s, s_{2}) ,
$$
where $b$ is the binomial probability. The rate is therefore at most $\alpha$ for every $\theta$ whenever $\mathrm{CRP}$ is at most $\alpha$ for every outcome. The column `CRP.total` is the rejection probability given only the total $s + s_{2}$, with the responder count of group 1 treated as one hypergeometric count among all final patients. Fisher's exact test keeps it at or below $\alpha$ by construction.

```{r crp}
crp <- BinaryCondRejectBSSR(
  Delta.A = 0.3, N1 = 12, N2 = 12, n.interim = c(6, 6), r = 1, alpha = 0.025,
  tar.power = 0.8, Test = 'Fisher', ss.method = 'standard', theta = c(0.3, 0.5)
)
crp
plot(crp)
```

`CRP` exceeds the nominal level at `r attr(crp, 'exceed')[['CRP']]` outcomes and `CRP.total` at `r attr(crp, 'exceed')[['CRP.total']]`. The allocation is fixed within each stage, so the responder count of group 1 given $s$ and $s_{2}$ is the sum of two hypergeometric counts rather than one, and `CRP` differs from `CRP.total` even when the sample size is not re-estimated. Whether the outcomes above the level raise the type I error rate is seen from the decomposition by $s$, which `theta` requests.

```{r by-s}
by.s <- attr(crp, 'by.s')
head(by.s)
attr(crp, 'TIE')
plot(crp, what = 'by.s')
```

`TIE.s` is the type I error rate given $s$, and `contribution` is its product with the probability `prob.s` of $s$. The contributions add up to the type I error rate at each value of `theta`, which is `r paste(signif(attr(crp, 'TIE')$TIE, 4), collapse = ' and ')` here.

## Non-inferiority

With a non-inferiority margin, `BinaryTypeIErrorBSSR()` and `BinaryAlphaAdjBSSR()` evaluate the rate on the boundary of the null hypothesis, parametrized by the pooled response probability, see the `bbssr-non-inferiority` vignette. `BinaryCondRejectBSSR()` covers tests of superiority only, since on the boundary of a non-inferiority hypothesis the two response probabilities differ and the conditional distributions of the responder counts depend on them.

## References

Friede, T. and Kieser, M. (2004). Sample size recalculation for binary data in internal pilot study designs. *Pharmaceutical Statistics*, 3, 269-279.

Kieser, M. (2020). *Methods and Applications of Sample Size Calculation and Recalculation in Clinical Trials*. Springer.

Kieser, M. and Friede, T. (2000). Re-calculating the sample size in internal pilot study designs with control of the type I error rate. *Statistics in Medicine*, 19, 901-911.
