---
title: "Comparing Designs with BinaryGridBSSR"
author: "Gosuke Homma"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comparing Designs with BinaryGridBSSR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r load}
library(bbssr)
```

## One data frame for many designs

Choosing a design means comparing several candidates, for example tests, timings of the interim analysis or assumed effects, over the same scenarios. `BinaryGridBSSR()` takes the candidates as the rows of a data frame, evaluates each with `BinaryPowerBSSR()` at common true pooled response probabilities, and returns one data frame with a row for each design and each probability.

The columns of the design data frame are arguments of `BinaryPowerBSSR()`. Arguments shared by all designs are given through `...`, and an argument may not appear in both places. Below, two tests are crossed with two timings of the interim analysis.

```{r grid}
design <- expand.grid(Test = c('Chisq', 'Fisher'), omega = c(0.3, 0.5),
                      stringsAsFactors = FALSE)
design
grid <- BinaryGridBSSR(design, p = seq(0.2, 0.5, by = 0.1), Delta.A = 0.3,
                       N1 = 40, N2 = 40, r = 1, alpha = 0.025, tar.power = 0.8,
                       ss.method = 'standard')
head(grid)
```

Each row holds the index `design`, the columns of the design data frame, the initial and interim sample sizes, the true effect, the true response probabilities, the power of the re-estimation design and of the fixed-sample design, and the expected value, standard deviation and quartiles of the final total sample size. `P.N.max` is the probability that the final total reaches the largest value the design can reach.

## Summaries and plots

`summary()` gives one row per design, with the range of the power over the scenarios and the largest expected total sample size.

```{r summary}
summary(grid)
```

`plot()` draws the power or the expected total sample size against the true pooled response probability, with one colour per value of a column and, optionally, one panel per value of another.

```{r plot-power}
plot(grid, colour.by = 'Test', facet.by = 'omega')
```

```{r plot-size}
plot(grid, what = 'E.N', colour.by = 'Test', facet.by = 'omega')
```

## Initial sample sizes from a planning proportion

Instead of the columns `N1` and `N2`, a design can give a pooled planning proportion `p.plan`. The planning proportions of the two groups are then obtained from `p.plan` and the assumed effect, as in the re-estimation, and the initial sample sizes are those of `BinarySampleSize()` with the test, level, target power and allocation ratio of the design, the method `ss.method` of the re-estimation and its rounding rule. With `type1 = TRUE` the largest type I error rates of each design and of its fixed-sample counterpart are added to the attribute `designs`, as computed by `BinaryTypeIErrorBSSR()`.

```{r planned}
planned <- data.frame(Delta.A = c(0.2, 0.25, 0.3), p.plan = 0.35)
grid.plan <- BinaryGridBSSR(planned, p = seq(0.2, 0.5, by = 0.1), omega = 0.5, r = 1,
                            alpha = 0.025, tar.power = 0.8, Test = 'Chisq',
                            ss.method = 'standard', type1 = TRUE)
attr(grid.plan, 'designs')
```

## Designs specified in different ways

A missing value in a column means that the argument is not given for that design. Designs whose interim analysis is given by `omega` and designs whose interim sample sizes are given by the columns `n1.interim` and `n2.interim` can therefore share one data frame.

```{r mixed}
mixed <- data.frame(omega = c(0.5, NA), n1.interim = c(NA, 12), n2.interim = c(NA, 12))
grid.mixed <- BinaryGridBSSR(mixed, p = 0.35, Delta.A = 0.3, N1 = 40, N2 = 40, r = 1,
                             alpha = 0.025, tar.power = 0.8, Test = 'Chisq',
                             ss.method = 'standard')
grid.mixed[, c('design', 'n1.interim', 'n2.interim', 'power.BSSR', 'E.N')]
```

## A true effect other than the assumed one

The power is evaluated at the assumed effect of each design unless `Delta.T` gives the true effect, either as an argument for all designs or as a column of the design data frame.

```{r true-effect}
grid.smaller <- BinaryGridBSSR(design, p = seq(0.2, 0.5, by = 0.1), Delta.T = 0.2,
                               Delta.A = 0.3, N1 = 40, N2 = 40, r = 1, alpha = 0.025,
                               tar.power = 0.8, ss.method = 'standard')
summary(grid.smaller)
```

Errors and warnings raised while a design is evaluated are prefixed with the index of the design, and `verbose = TRUE` reports the progress. The p-values of each test are kept for the rest of the session, so designs that share sample sizes are evaluated faster after the first one.
